Design and implemention of open health data systems with standardized data and analytics capabilities for low- and middle income countries

### 2024-08-26 index.qmd
wip

---
title: "Conceptualizing and implementing open, decentralized health data sharing platforms for low- and middle income countries"
format:
  elsevier-html:
    toc: true
    toc-depth: 3
    fig-cap-location: top
    code-fold: true
  pdf: default
    # keep-tex: true
    # journal:
    #   name: JIMR
    #   cite-style: authoryear
author:
  - name: Daniel Kapitan
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
      - name: Eindhoven University of Technology
        city: Eindhoven
        country: the Netherlands
    orcid: 0000-0001-8979-9194
    email: daniel@kapitan.net
  - name: Femke Heddema
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Julie Fleischer
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Chris Ihure
    affiliations:
      - name: PharmAccess Kenya
        city: Nairobi
        country: Kenya
  - name: Steven Wanyee
    affiliations:
      - name: IntelliSOFT
        city: Nairobi
        country: Kenya
  - name: Alessandro Pietrobon 
    affiliations:
      - name: ONA
        city: Nairobi
        country: Kenya
  - name: Ryan Chrichton
    affiliations:
      - name: Jembi Health Systems
        city: Durban
        country: South Africa
  - name: John Grimes
    affiliations:
      - name: Australian e-Health Research Centre
        city: Brisbane
        country: Australia
  - name: Paula van Brakel
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands

  - name: Mark van der Graaf
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Nicole Spieker
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands

@@

Digital transformation have the potential to make inclusive health systems in low- and middle income countries (LMICs) a reality by connecting patients and professionals, using data as a currency of trust. Adoption of digital technologies in healthcare means more than simply increasing efficiency or cutting costs. Instead, digital technologies have the potential to i) increase the availability, accessibility, acceptability, and quality of health services as we know them; ii) change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile through the use of such technologies; and iii) enfranchise patients and communities (and particularly vulnerable groups including children and young people) and evolve their relationship with health professionals and providers, thus helping shape the health system according to the needs of the patients and communities [@kickbusch2021lancet].

@@

To achieve scaleability of digital health interventions, we need to conceptualize and implement digital platforms and their associated ecosystems that constitute an important enabling socio-technical infrastructure upon which such interventions can be realized in a more standardized and sustainable fashion. Digital platforms are a challenging research object because of their distributed nature and intertwinement with institutions, markets and technologies [@dereuver2018digital]. Digital platform ecosystems vary according to three core building blocks: (1) platform ownership, (2) value-creating mechanisms, and (3) complementor autonomy [@hein2020digital]. Various studies have investigated why the healthcare domain has been slow in adopting platforms and their ecosystems [@hermes2020digital;@ozalp2022digital]. Before presenting our main contributions, we highlight two paradigmatic themes that are relevant to our paper.

### From centralized to decentralized platforms

Digital platforms have predominantly flourished in a business setting, as is exemplified by the well known success stories of cloud platforms providers (AWS, Microsoft, Google, Alibaba), two-sided market platforms (Airbnb, Spotify, Uber) and social media platforms (Meta, WeChat). While these examples allude to the potentional to generate social and economic value by using platforms as a key enabler of digital transformation of healthcare in LMICs, it is less obvious in what way a digital platform can trigger specific positive effects *for* development [@bonina2021digital]. Part of the problem in grasping the implications of platforms for development in the field of information systems has been the lack of clarity regarding the understanding of what digital platforms are, what their main features are and how they generate value [@dereuver2018digital]. Notably, even valuable landscape papers in information systems largely ignore the development implications of digital platforms [@dereuver2018digital] or may summarize them in rather naïve ways [@bonina2021digital]. Put more strongly, issues pertaining to the winner-takes-all nature of digital platforms have resulted in so-called 'transgression of spheres' where commercial platform providers acquire disproprotionately large market power which is at odds with realizing public value for the common good [@sharon2018when].

It almost seems we have forgotten that digital platforms need not always be centralized, but that we have a choice to implement decentralized or distributed platforms. In his seminal paper in 1963, Baran distinguished three types of network topologies ([@fig-network-topologies]). In a centralized network, all communication is facilitated through a single, central hub, which is obviously vulnerable as "...destruction of single node destroys communication between the end stations" [@baran1964distributed]. At the opposite end of the spectrum, in a fully distributed network, stations are connected to all adjacent stations. In practice, a mixture of centralized and distributed components is the most commonly used topology, aptly called decentralized networks as they remove complete reliance upon a single point, while at the same time where is a hierarchy where smaller centralized networks are connected to higher-lying networks through interconnecting hubs (encircled in blue). Note that the internet has both the characteristics of a distributed network (for example, traffic and data processing are shared across multiple paths and servers) as well as that of a decentralized network (there is no single entity that controls the entire internet. Instead, numerous internet service providers (ISPs), organizations, and governments manage different parts of the network).

![Three network topologies, taken from Baran (1964). The nodes encircled in blue in the decentralized network are hubs that provide connections between sub-networks.](images/network-topologies.png){#fig-network-topologies}

We are inspired by various fields of study that have shown that decentralized and/or distributed systems are often 'resilient by design' [@fiksel2003designing;@liu2010architectural;@mayar2022resilience], for example, allowing efficient deployments at different geographical scales. Given that LMICs are resource-constrained, we consider downward scaleability as a key design criterium [@fayad2005scalable;@duboc2006framework]. We consider the insights from the 'hourglass model', which provides clear design principles how the use of minimal standards in a layered, distributed systems allows implementation information systetms using a great diversity of services [@beck2019hourglass].

@@

### From open source to open data sharing

Case studies on digital platforms in healthcare point to an emerging pattern where the focus shifts from the digital platform with its defining software and hardware components, to the data as the primary object of interest in and of itself [@ozalp2022digital;@alaimo2022organizations]. This observation ties into with the proposed research agenda by de Reuver et al. to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness]. @tbl-comparing-platforms compares the key characteristics of digital vs. data platforms. The concept of health data spaces, currently under development in Europe, is an archetypical example of a data platform [@otto2022designing;@shabani2022will].

| | Digital platforms (e.g. software, gaming) | data platforms (e.g. data marketplaces) |
|:---|:-----|:-----|
| **User groups** | Developers, Consumers | Data consumers, Data owners, Solution providers |
| **Object of openness** | Software/hardware modules | Data, aggregated data, trained models; Data analytics modules |
| **Ways to realize openness** | Open source, open boundary resources (APIs) | Data-driven mechanisms (secure computational environments) |
| **Market consolidation** | Strong network effects; Dominant cross-industry platforms | Moderate network effects; Fragmentation and heterogeneity |

: Comparing digital and health data platforms, taken from de Reuver et al. (2022) {#tbl-comparing-platforms .striped .hover .responsive}

To illustrate the relevance of conceptualising health platforms as data platforms, consider the OpenHIE framework [@openhie] which has been adopted by many sub-Saharan African countries as the architectural blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one]. The OpenHIE standard is best described as a digital platform as it is designed to facilitate information exchange, using specific workflows and information standards (open boundary resources in @tbl-comparing-platforms), with a primary focus is on creating a cohesive infrastructure that supports the interaction and integration of various health information systems. Since its inception in 2013, the ICT for Development community (ICT4D) has pushed for the development of open source implementations of OpenHIE components, which are now available as digital public goods [@digitalpublicgoods]. @tbl-digital-public-goods list the currently available open source components that adhere to the OpenHIE specification.

| Category  | Component | Digital Public Good |
|:----------|:----------|:--------------------|
| **Interoperability layer (IOL)** | IOL | [OpenHIM](http://openhim.org) |
| | | [mHero](https://www.mhero.org/technology) |
| | | [OpenFN](https://docs.openfn.org/) |
| **Registry Services** | Client Registry (CR) | [SanteMPI](https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi) |
| | | [JeMPI](https://jembi.gitbook.io/jempi/) |
| | | [OpenCR](https://www.openclientregistry.org/) |
| | | [OpenCRVS](https://www.opencrvs.org/) |
| | Facility Registry (FR) | [Global Open Facility Registry (GOFR)](https://github.com/intrahealth/gofr) |
| | | [GeoPrism Registry (GPR)](https://geoprismregistry.com/) |
| | | [Resource Map](http://instedd.org/technologies/resource-map/) |
| | | [Healthsite.io](https://healthsites.io/) |
| | | [GeoPrism Registry](https://geoprismregistry.com/) |
| | | [DHIS2](https://github.com/dhis2) |
| | Health Worker Registry (HWR) | [iHRIS](https://www.ihris.org/ihris-50) |
| | Terminology Service (TS) | [OCL Terminology Service](https://openconceptlab.org/terminology-service/) |
| | Product Catalog (PC) | [PCMT](https://productcatalog.io/) |
| **Business Domain Services** | Shared Health Record (SHR) | [HAPI FHIR](https://hapifhir.io/) |
| | | [Fhirbase](https://www.health-samurai.io/fhirbase) |
| | | [FHIR Server for Azure](https://github.com/microsoft/fhir-server) |
| | | [I-TECH-UW SHR](https://github.com/I-TECH-UW/shared-health-record) |
| | Health Management Information System (HMIS)[^1] | [DHIS2](http://dhis2.org/) |
| | Finance and Insurance Service (FIS) | [OpenIMIS](http://openimis.org/) |
| | Logistics Management Information System (LMIS) | [OpenLMIS](http://openlmis.org/) |
| | | [OpenBoxes](https://openboxes.com/features/) |
| **Generic** | Identity and Access Management (IAM) | [Keycloak](https://www.keycloak.org/) |
| | Gateway & proxy | [FHIR Information Gateway](https://github.com/google/fhir-gateway) |
| | Admin dashboard for SHR | [OpenSRP FHIR Web](https://github.com/onaio/fhir-web) |
| | Configuration and deployment | [Instant OpenHIE](https://jembi.gitbook.io/instant-v2/) |

: Overview of current open source implementations of components that fit in the OpenHIE specification. Point-of-Service systems are excluded for brevity. List compiled using [OpenHIE Reference Technologies](https://wiki.ohie.org/display/DR/OpenHIE+Reference+Technologies), [Global Goods for Digital Health](https://digitalsquare.org/digital-health-global-goods), [Digital Public Goods Alliance](https://digitalpublicgoods.net/) and search of open source code repositories. A systematic review of such digital public goods is beyond the scope of this paper.  {#tbl-digital-public-goods .striped .hover .responsive}

[^1]: Note that there is often a confusion on the acronym HMIS. Strictly speaking, the OpenHIE specification uses HMIS to refer to a Health Management Information System that is part of the Business Domain Services. Sometimes HMIS is used to refer to a Hospital Management Systems in the Point-of-Service domain, synonymous with an Electronic Medical Record (EMR) system.

This shift in perspective also has real implications for the design and implementation of OpenHIE. We observe that some African countries have, as a matter of course, extended the OpenHIE framework to include “data & analytics services” as an additional domain in their policies [@mbugua2021adoption;@dalhatu2023paper]. The rationale for this addition is indeed to facilitate secondary reuse of health data for academic research, real-world evidence studies etc. using the principles of Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair]. In doing so, however, we have implicitly moved from conceptualizing digital health platforms for primary data sharing (the original OpenHIE specification) to health data platforms for secondary data sharing. If we are to use the OpenHIE framework as an open data platform as well, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs. One of the motivations for this paper is integrate recent standards and open source implementations from the data engineering community into the OpenHIE framework.

we seek to specifiy and demonstrate how this can be implemented using the full-STAC approach [@mehl2023fullstac], that consists of:

1. FHIR as the main open standard
2. the Instant HIE framework for integrating open technologies
3. an extended version of OpenHIE as the open architecture that includes a more detailed architecture for the data platform
4. open content with various demonstrators in the domain of antenatal and maternal care (ANC) as a first version of hands-on digital adaptation kits.

#### 1. Sharing of data sets at the most granular (patient) level, persisted as longitudinal records
The Shared Health Record (SHR) in OpenHIE provides an operational, real-time transactional data source which is intended for primary data sharing. The specification explicitly states that the SHR is distinct from a datawarehouse. At the same time, the Health Management Information System (HMIS) component is specified to fulfil functional requirements that are typically provided by datawarehouses.

- **Q1:** How can proven solution designs and modern technologies from data warehousing and engineering be integrated in the OpenHIE specification and how does this relate to the curren specifications of the SHR and HMIS.

#### 2. Sharing of data products derived from the original data set
OpenHIE specifies that the Health Management Information System (HMIS) component should support a workflow to validate and save aggregate data based on the emerging [IHE Aggregate Data Exchange (ADX) standard](https://www.ihe.net/uploadedFiles/Documents/QRPH/IHE_QRPH_Suppl_ADX.pdf). This is just one specific workflow within a larger data analytics value chain that covers i) collection, ii) standardization, iii) cleaning, iv) storage, v) analysis and vi) distribution [@curry2016big]. Sharing of benchmarking, and decision-support tools based on various statistical (machine learning) models that have been trained on data sets and/or data product, can be shared for standalone use.

- **Q2:** How can proven solutions designs from the data warehousing and engineering community be included in OpenHIE to support sharing of wider variety of data products.

#### 3. Sharing of secure computational environments to access and work with the data

Current solution design of centralized data storage is increasingly being challenged. Recent work points to federated learning (FL) [@rieke2020future] and privacy- enhancing technologies (PETs) [@scheibner2021revolutionizing;@jordan2022selecting] as a better long-term solution to secure and equitable sharing of data. Given the resource constrained environments of LMICs, is is particularly challenging to devise a solution that is economically feasible and ...

## Methods

In this paper, we present a design that extends the OpenHIE specification to include the three types of data sharing mentioned above. Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular OHDP that can be configured and reused across a variety of use-cases. We employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds]. Given that it is prohibitively expensive to evaluate the design a real-world setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three examples of health data platforms in LMICs, namely:

 1. the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/))
 2. the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/)
 3. the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).

These three real-world implementations are evaluated along the following dimensions:

- What is the level of openness of the implementation, specifically in terms of the three types of data sharing and platform-to-platform opennenss?
- How are the core maintenance functions of the OHDP implemented, where we particularly focus on functions related to metadata management, data lineage and access control mechanisms?
- Is the solution suitable for downward scalability, where we focus on the computer and storage requirements of the solution and assess the minimal requirements for on-site resources (computers, edge devices)

As part of our design research, we have taken a narrative approach in surveying existing scientific studies on health data platforms, focusing on the seminal reports and subsequently searching forward citations. In addition, we have searched the open source repositories (most notably GitHub) and the online communities (OpenHIE community, FHIR community) to search for relevant open standards, technologies and architectures. This paper should not be considered as a proper systematic review.

## Results

### Design

#### Open standards: using FHIR as the common data model

The recent convergence to FHIR as the de facto standard for information exchange has fuelled the development of OpenHIE. FHIR is currently used both for routine healthcare settings [@amar2024electronic;@ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well. The guidelines and standards of the African Union explicitly state FHIR is to be used as the messaging standard [@2023african]. The FHIR-native OpenSRP platform [@mehl2020open] has been deployed in 14 countries targeting various patient populations, amongst which a reference implementation of the WHO antenatal and neonatal care guidelines for midwives in Lombok, Indonesia [@summitinstitutefordevelopment2023bunda;@kurniawan2019midwife]. In India, FHIR is used as the underlying technology for the open Health Claims Exchange protocol specification, which has been adopted by the Indian government as the standard for e-claims handling [@hcx]. This range of utilizations showcase the standards’ widespread applicability. The proceedings of the OpenHIE conference 2023 attest to the fact that FHIR and open source technologies are embraced as critical enablers in implementing health information exchanges in LMICs [@ohie2023unconference].

Various studies have investigated the merits of FHIR and its performance vis-a-vis other healthcare standards. Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data;@tsafnat2024converge]. A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR concluded that i) these three standards are functionally and technically compatible, and therefore can be used side by side; and that ii) each of these standards have their strengths and limitations that correlate with their intended use as summarized in the @tbl-comparison.

![Comparison of OpenEHR, ISO 13606 and FHIR standards](images/comparison-ehr-standards.png){width=auto #tbl-comparison}

For an infectious diseases dataset with a limited scope, OpenEHR, OMOP and FHIR have been compared and found all to be equally suitable [@rinaldi2021openehr]. Comparing OMOP and FHIR, the latter has been found to support more granular mappings required for analytics and was therefore chosen as the standard for the CODA project [@mullie2023coda].

Although FHIR was originally designed only for exchange between systems, we propose to use it as the common data model for the design presented here for the following reasons:

- Industry adoption has significantly increased, as exemplified by FHIR-based offering by major cloud providers such as Google, Azure and AWS. Also, Africa CDC has explicitly chosen FHIR as the preferred standard;
- The widespread availibility of the Bulk FHIR API [@mandl2020push;@jones2021landscape] enables bulk, file-based batchwise processing for analytics using the lakehouse architecture as detailed in the next section;
- The concept of FHIR Profiles allow localisation to tailor the standard to a specific use case. A profile defines rules, extensions, and constraints for a resource. We posit that the possible penalty of this flexibility, namely having to manage different FHIR versions and/or profiles, is less of an issue in the context of LMICs where first priority is to exchange datasets such as the International Patient Summary (IPS) that are less complex compared to the requirements for high income countries;
- Being based on webstandards, the FHIR standard lends itself best for further separation of concerns as envisioned by the composable data stack. This is an important enabler for the downward scalability of the solution;
- With its inherent, graph-like nature, FHIR can be readily incorporated into the principles of FAIR data sharing, where FHIR-based data repositories can be integrated in an overarching netwerk of FAIR data stations [@sinaci2023data;@pedrera-jimenez2023can].

TO DO:
- tie-in paper [@tsafnat2024converge] on why we use FHIR
- re-use these elements from table 1 into this section:

  - Given that the Bulk FHIR API has by now been incorporated in all major FHIR implementations [@mandl2020push;@jones2021landscape], the FHIR standard can be readily used to bridge the gap between primary (transactional) and secondary (analytics) data sharing.
  - __SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows. Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items. Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].  Modern data platforms take this Within the HL7 ecosystem much progress has been made to support "analytics-on-FHIR" with standards including [FHIRPath](https://hl7.org/fhirpath/) and the new SQL-on-FHIR specification [@sql-on-fhir]. Given the use of FHIR as the common data model (the rationale of which will be described later), we formulate the second design question:

Possible risks pertaining to the use of FHIR as the common data model, most notably the possible incompatibilities and/or high costs of maintenance in supporting different versions, will be addressed in the Discussion.

#### Open architecture: extending OpenHIE with a composable data stack

Data management and analytics platforms have undergone significant changes since the first generation of data warehouses were introduced. Recent studies have shown that the current practice has converged towards the lakehouse as one of the most commonly used solution designs [@armbrust2021lakehouse;@hai2023data;@harby2022data]. Lakehouses typically have a zonal architecture that follow the Extract-Load-Transform pattern (ELT) where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T) [@hai2023data]. The discerning characteristic of the lakehouse architecture is its foundation on low-cost and directly-accessible storage that also provides traditional database management and performance features such as ACID transactions, data versioning, auditing, indexing, caching, and query optimization [@armbrust2021lakehouse]. Lakehouses thus combine the key benefits of data lakes and data warehouses: low-cost storage in an open format accessible by a variety of systems from the former, and powerful management and optimization features from the latter.

With respect to current implementations of lakehouse data platforms, we observe a proliferation of tools with as yet limited standards to improve technical interoperability. In the analysis of Pedreira et al. [@pedreira2023composable] the requirement for specialization in data management systems has evolved faster than our software development practices. This situation has created a siloed landscape composed of hundreds of products developed and maintained as monoliths, with limited reuse between systems. It has also affected the end users, who are often required to learn the idiosyncrasies of dozens of incompatible SQL and non-SQL API dialects, and settle for systems with incomplete functionality and inconsistent semantics. To remedy this, Pedreira et al. call to (re-)design and implement modern data platforms in terms of a 'composable data stack’ as a means to decrease development and maintenance cost and pick-up the speed of innovation.

While the lakehouse architecture separates the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further. A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A, IR) and Apache Arrow for data connectivity and data memory layout (standards B and C, respectively). The first generation of open source components are already available. For example, the Ibis user interface is currently sufficiently mature to offer a standardized dataframe interface to 19 different execution engines [@ibisproject].

::: {#fig-composable-data-stack layout-nrow=2 fig-cap="Schematic overview of the composable data stack showing a) the overall architecture and b) examples of implementations. Images taken from [https://voltrondata.com/codex](https://voltrondata.com/codex)."}
![a)](images/composable-data-stack.png)

![b)](images/composable-data-stack-implementation-2.png)
:::

@@

#### Extending OpenHIE with a FHIR-based composable data stack

We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different service layers by synthesizing the current best practices of the a lakehouse architecture of [@hai2023data;@harby2022data;@harby2024data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services). These 5 services are considered the core of the FHIR data platform, and although in practice often a downstream dashboarding or visualization component is used, that component is not the main focus of our analysis. Rather, we aim to elucidate and conceptualize the core "FHIR Data Lakehouse".

+-------------------+---------------------------------------------------------------------------------------+
| Service           | Functional requirements                                                               |
+===================+=======================================================================================+
| Ingestion         | - Bulk                                                                                |
|                   | - Streaming                                                                           |
+-------------------+---------------------------------------------------------------------------------------+
| Storage           | - File-based blob storage                                                             |
|                   | - Database optimized for online analytical processing (OLAP)
+-------------------+---------------------------------------------------------------------------------------+
| Maintenance       | - SQL-on-FHIR View defintions                                                         |
|                   | - Catalog and other maintenance-related functions as defined by Hai et al.            |
+-------------------+---------------------------------------------------------------------------------------+
| Processing        | - SQL-on-FHIR Runner                                                                  |
| & API             | - Execution engine on tabular data as defined in composable data stack                |
|                   | - Capability to participate as a node in federated learning / MPC network             |
|                   | - Read-only access to storage                                                         |  
+-------------------+---------------------------------------------------------------------------------------+
| Data              | - SQL interactive development environment (IDE)                                       |
| Consumer          | - Interactive notebook computing environment [@granger2021jupyter]                    |
|                   | - BI reporting, dashboarding and data visualization                                   |  
+-------------------+---------------------------------------------------------------------------------------+

: Definition of Data and Analysis Services {#tbl-data-and-analysis-services}

@@

### Open content

...

### Evaluation of implementations

#### Jembi OpenHIM platform

#### ONA OpenSRP 2

#### Level of opennness

TO DO: evaluate openness of OpenHIM platform

#### Core maintenance functions

TO DO: evaluate maintenance functions

#### Downward scaleability

TO DO: evaluate downward scalability

@@

DK received funding from PharmAccess to conduct this work as a contractor.
SW/IntelliSOFT received funding from PharmAccess as an implementation partner in various projects.
AP/ONA received funding from PharmAccess to develop a new version of components of the open source OpenSRP 2 framework.

>>>>

---
title: "Design and implemention of open health data systems with standardized data and analytics capabilities for low- and middle income countries"
format:
  elsevier-html:
    toc: true
    toc-depth: 3
    fig-cap-location: top
    code-fold: true
  pdf: default
    # keep-tex: true
    # journal:
    #   name: JIMR
    #   cite-style: authoryear
author:
  - name: Daniel Kapitan
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
      - name: Eindhoven University of Technology
        city: Eindhoven
        country: the Netherlands
    orcid: 0000-0001-8979-9194
    email: daniel@kapitan.net
  - name: Femke Heddema
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Julie Fleischer
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Irni Gemzon
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Chris Ihure
    affiliations:
      - name: PharmAccess Kenya
        city: Nairobi
        country: Kenya
  - name: Steven Wanyee
    affiliations:
      - name: IntelliSOFT
        city: Nairobi
        country: Kenya
  - name: Alessandro Pietrobon 
    affiliations:
      - name: ONA
        city: Nairobi
        country: Kenya
  - name: Ryan Chrichton
    affiliations:
      - name: Jembi Health Systems
        city: Durban
        country: South Africa
  - name: John Grimes
    affiliations:
      - name: Australian e-Health Research Centre
        city: Brisbane
        country: Australia
  # - name: Paula van Brakel
  #   affiliations:
  #     - name: PharmAccess Foundation
  #       city: Amsterdam
  #       country: the Netherlands
  # - name: Mark van der Graaf
  #   affiliations:
  #     - name: PharmAccess Foundation
  #       city: Amsterdam
  #       country: the Netherlands
  # - name: Nicole Spieker
  #   affiliations:
  #     - name: PharmAccess Foundation
  #       city: Amsterdam
  #       country: the Netherlands

@@

Digital transformation have the potential to make inclusive health systems in low- and middle income countries (LMICs) a reality by connecting patients and professionals, using data as a currency of trust. Digital technologies have the potential to increase the availability, accessibility, acceptability, and quality of health services as we know them; change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile through the use of such technologies; and enfranchise patients and communities (and particularly vulnerable groups including children and young people) and evolve their relationship with health professionals and providers, thus helping shape the health system according to the needs of the patients and communities [@kickbusch2021lancet].

@@

To achieve scaleability of digital health interventions, we need to design and implement standardized open health data systems (OHDS) and their associated ecosystems that can support the wider health sector in improving along five dimensions, namely [@kelley2020digital]:

1. overall quality and continuity of care;
2. adherence to clinical guidelines and best practices;
3. efficiency and affordability of services and health commodities, by reducing duplication of effort and ensuring effective use of time and resources;
4. health-financing models and processes, regulation, oversight, and patient safety resulting from increased availability of performance data and reductions in errors; and
5. health policy-making and resource allocation based on better quality data .

### Data & analytics functions are essential in open health data systems

Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions is indeed within reach. As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs. By actively coaxing pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of MCH services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].

MomCare, like most other digital health interventions, is critically dependent on the availability of data and analytics functions within the OHDS. For example, the widely used DHIS2 reports that are most commonly used in LMICs contained insufficient information to run the programme. More detailed analytics at the patient level were required. It is, in fact, a widely held belief that data & analytics are an essential part of OHDSs. This is exemplified how the OpenHIE reference architecture [@openhie] has been adopted by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one]. These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain [@mbugua2021adoption;@dalhatu2023paper]. The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc. which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].

## Extending OpenHIE to include modern data and analytics standards

If we intend to use the OpenHIE standard not only for primary health data sharing, but also to support data and analytics for secondary health data sharing, we need to extend its specification. The main contribution of this paper is to propose how recent standards and open source implementations from the data engineering community could be integrated into the OpenHIE framework, as illustrated in @fig-ohie.

The original OpenHIE specification discerns four domains, namely Point-of-Service systems, the Interoperability Layer, Common Services and Business Services. We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different zones taken from the data lakehouse architecture, which currently is the most commonly used design pattern in this domain [@armbrust2021lakehouse;@hai2023data;@harby2022data;@harby2024data]. Lakehouses typically have a zonal architecture that follow the Extract-Load-Transform pattern (ELT) where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T) [@hai2023data]. The discerning characteristic of the lakehouse architecture is its foundation on low-cost and directly-accessible storage that also provides traditional database management and performance features such as ACID transactions, data versioning, auditing, indexing, caching, and query optimization [@armbrust2021lakehouse]. Lakehouses thus combine the key benefits of data lakes and data warehouses: low-cost storage in an open format accessible by a variety of systems from the former, and powerful management and optimization features from the latter.

Following the terminology proposed by Hai et al. [@hai2023data], @tbl-data-and-analysis-services summarizes how the different zones of the lakehouse architecture can be adapted for healthcare and integrated into the OpeHIE specification. Taking Fast Healthcare Interoperability Resources (FHIR) as the open data standard, we envisage the extended OpenHIE architecture to include a 'FHIR lakehouse'[^1].

[^1]: We have argued the choice of FHIR as the common data model elsewhere, working paper to be submitted ([link](https://plugin-healthcare.github.io/viewpoint-jimr/)).

+-------------------+---------------------------------------------------------------------------------------+
| Service           | Functional requirements                                                               |
+===================+=======================================================================================+
| Ingestion         | - Bulk                                                                                |
|                   | - Streaming                                                                           |
+-------------------+---------------------------------------------------------------------------------------+
| Storage           | - File-based blob storage                                                             |
|                   | - Database optimized for online analytical processing (OLAP)
+-------------------+---------------------------------------------------------------------------------------+
| Maintenance       | - SQL-on-FHIR View defintions                                                         |
|                   | - Catalog and other maintenance-related functions as defined by Hai et al.            |
+-------------------+---------------------------------------------------------------------------------------+
| Processing        | - SQL-on-FHIR Runner                                                                  |
| & API             | - Execution engine on tabular data as defined in composable data stack                |
|                   | - Capability to participate as a node in federated learning / MPC network             |
|                   | - Read-only access to storage                                                         |  
+-------------------+---------------------------------------------------------------------------------------+
| Data              | - SQL interactive development environment (IDE)                                       |
| Consumer          | - Interactive notebook computing environment [@granger2021jupyter]                    |
|                   | - BI reporting, dashboarding and data visualization                                   |  
+-------------------+---------------------------------------------------------------------------------------+
: Adaptation of lakehouse zones for healthcare specific standardization of Data and Analysis Services {#tbl-data-and-analysis-services}

[TO DO: need to explain table (is too condensed now) and especially FHIR specific components.]

In the remainder of this paper, we first describe in more detail how the FHIR lakehouse pattern can be implemented using existing open source technologies. We do so by presenting work done at PharmAccess. Code and digital artifacts are made available [TO DO: include links to support GitHub repositories].

Subquently, we compare this solution design with two widely used OpenHIE-compliant frameworks, namely the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)) and the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/). We take a narrative approach in presenting our design and the case studies, surveying existing scientific studies on OHDSs, focusing on the seminal reports and subsequently searching forward citations. In addition, we have searched the open source repositories (most notably GitHub) and the online communities (OpenHIE community, FHIR community) to search for relevant open standards, technologies and architectures. This paper should not be considered as a proper systematic review.

Finally, we discuss our findings and propose routes for future development.

## Case study of implementations

### Jembi OpenHIM platform

### ONA OpenSRP 2

@@

| Level | Description | Number of facilities | 
|:---:|----------------------|--------:|
| 2 | Dispensaries and private clincs, typically located in a school, industrial plant or other organization that dispenses medication and sometimes basic medical and dental treatment | 8,806 |
| 3 | Health centres, medium-sized units which cater for a population of about 80,000 people | 2,559 |
| 4 | Sub-county hospital, similar to health centres with additional facilities for more complex procedures |   971 |
| 5 | County referral hospital, regional centres which provide specialised care |    34 |
| 6 | National referral hospital | 5 |

: Number of healthcare facilities in Kenya. Source: Kenya Health Facility Census, Ministry of Health, September 2023. {#tbl-kenya-facilities}

@@

@@

@@

DK received funding from PharmAccess to conduct this work as a contractor.
SW/IntelliSOFT received funding from PharmAccess as an implementation partner in various projects.
AP/ONA received funding from PharmAccess to develop an improved version of one of the components of the open source OpenSRP 2 framework.
